### 2025-11-15 sections/1-introduction.tex
brought back some stuff

Given a distance function on a dataset, $k$-nearest neighbors ($k$-NN) search returns the $k$ closest points to a query. We call algorithms with imperfect recall \textit{approximate} and those matching brute-force recall \textit{exact}. Most fast methods are approximate~\cite{gao2023high}; while sufficient for some applications, there remains demand for \textit{fast exact} search~\cite{ukey2023survey}.

@@

Modern k-NN search strategies include entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}, Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, and random projection and tree building (ANNOY)~\cite{annoy}.
% HNSW~\cite{malkov2016hnsw} relies on navigable small world (NSW) networks and skip lists.
% FAISS-IVF~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches only within cells near the query.
Crucially, many of these algorithms do not support \emph{exact} search.

@@

CAKES is a set of three such algorithms for exact $k$-NN search: Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
In the next section, we describe the clustering and search algorithms, and provide a complexity analysis.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a random uniform distribution.
Since CAKES is designed for datasets with a manifold structure, an analysis assuming a uniform distribution would be uninformative.
Therefore, our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.

>>>>

Given a distance function on a dataset, $k$-nearest neighbors ($k$-NN) search finds the $k$ closest points to a query.
We call algorithms with imperfect recall \textit{approximate} and those matching brute-force \textit{exact}.
Most fast methods are approximate~\cite{gao2023high}, and while they are sufficient for some applications, there remains a need for \textit{fast exact} search~\cite{ukey2023survey}.

@@

Modern k-NN search strategies include entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered} which leverages manifold structure, Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw} which relies on NSW networks and skip lists, InVerted File indexing (FAISS-IVF)~\cite{faissivf} which creates high-dimensional Voronoi cells, and ANNOY~\cite{annoy}, which builds several trees under different random projections.
Crucially, many of these algorithms do not support \emph{exact} search.

@@

CAKES is a set of three such algorithms for exact $k$-NN search: Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
In the next section, we describe the clustering and search algorithms, and provide a complexity analysis.
Our analysis assumes datasets with manifold structure, reflecting the conditions for which CAKES is designed.
We expect, and show in Section~\ref{sec:results}, that CAKES performs well on such datasets, and poorly when manifold structure is absent.

### 2025-11-15 sections/2-methods.tex
brought back some stuff

The manifold hypothesis~\cite{fefferman2016testing} states that high($D$)-dimensional data collected from constrained generating phenomena occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property:
\begin{equation}
    \text{LFD}(p, r_1, r_2) = \frac{\text{log} \big( \frac{|B(p, r_1)|}{|B(p, r_2)|} \big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $p$ is a dataset point and $B(p, r)$ is the set of points within radius $r$ of $p$.
When $d \ll D$ for the majority of points in the dataset, we say the dataset has low LFD.

@@

In the following, we use $\mathcal{Q}$ for the data structure containing clusters that are in contention for containing the $k$ nearest neighbors and $\mathcal{H}$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
We assume that the query $q$ is drawn from the same distribution as $\textbf{X}$. 
For a cluster $\mathcal{C}$, we denote by $\delta_\mathcal{C}$, $\delta^{+}_\mathcal{C}$, and $\delta^{-}_\mathcal{C}$, the distances from $q$ to the cluster center, to the theoretically farthest point, and to the theoretically closest point, respectively.

\subsubsection{Repeated \texorpdfstring{$\rho$}{p}-NN}
\label{sec:methods:knn-search:repeated-rnn}

We begin by using \textsc{TreeSearch} (Alg~\ref{alg:methods:rnn-search:tree-search}) on the root cluster $\mathcal{R}$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|\textbf{X}|$.
If the returned set $\mathcal{Q}$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $\mathcal{Q}$ is non-empty.
Now, so long as $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}|}} \big)^\frac{1}{\mu}$,
where $\mu$ is the harmonic mean of the LFD of the clusters in $\mathcal{Q}$, to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low, because in these cases, we can afford to increase the radius quickly. It will be small when we have found almost all $k$ points or LFD is large, and thus we want to increase the radius slowly.
Once $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| \geq k$, we perform \textsc{LeafSearch} (Alg~\ref{alg:methods:rnn-search:leaf-search}) to find the $k$ nearest neighbors.
\begin{algorithm}[h]
    \caption{\textsc{TreeSearch}($\mathcal{C}$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $\mathcal{C}$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_\mathcal{C} \leq r$}
            \State \textbf{return} $\{\mathcal{C}\}$
        \EndIf
        \State $[\mathcal{L}, \mathcal{R}]$ $\Leftarrow$ \textit{children} of $\mathcal{C}$
        \State \textbf{return} \Call{TreeSearch}{$\mathcal{L}, q, r$} $\cup$ \Call{TreeSearch}{$\mathcal{R}, q, r$}
    \end{algorithmic}
\end{algorithm}
\vspace{-0.2in}

@@

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(\textbf{X})}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\mathcal{N}_{\hat{r}}(\textbf{X})$ is the minimum number of clusters of radius $\hat{r}$ needed to cover $\textbf{X}$, and $\rho$ is the search radius.

>>>>

The manifold hypothesis~\cite{fefferman2016testing} states that high($D$)-dimensional data collected from constrained generating phenomena occupy a low($d$)-dimensional manifold within their embedding space ($d \ll D$).
We define the \emph{local fractal dimension} (LFD) to quantify this property:
\begin{equation}
    \text{LFD}(p, r_1, r_2) = \frac{\text{log} \big( \frac{|B(p, r_1)|}{|B(p, r_2)|} \big) }{\text{log} \big( \frac{r_1}{r_2} \big) }
    \label{eq:methods:lfd-original}
\end{equation}
where $p$ is a dataset point and $B(p, r)$ is the set of points within radius $r$ of $p$.
For a cluster $\mathcal{C}$ with center $c$ and radius $r$, we define its LFD as $\text{LFD}(c, r, r/2)$.
If the median LFD of clusters at most (and particularly high) depths in the tree is $\leq 3$, we say the dataset has ``low LFD''.

@@

In the following, we use $\mathcal{Q}$ for the data structure containing clusters that are in contention for containing the $k$ nearest neighbors and $\mathcal{H}$ for a max-priority queue (of size $k$) containing the current $k$ nearest neighbors found.
We assume that the query $q$ is drawn from the same distribution as $\textbf{X}$. 
For a cluster $\mathcal{C}$ with center $c$ and radius $r$, we define $\delta_\mathcal{C} = f(q, c)$, $\delta^{+}_\mathcal{C} = \delta_\mathcal{C} + r$, and $\delta^{-}_\mathcal{C} = \min(0, \delta_\mathcal{C} - r)$.

\subsubsection{Repeated \texorpdfstring{$\rho$}{p}-NN}
\label{sec:methods:knn-search:repeated-rnn}
This algorithm repeats parts of $\rho$-NN search~\cite{ishaq2019clustered}, which finds all points within a radius $\rho$ around a query.
Repeated $\rho$-NN begins by using \textsc{TreeSearch} (Alg~\ref{alg:methods:rnn-search:tree-search}) on the root cluster $\mathcal{R}$ with the search radius $r$ equal to the radius of $\mathcal{R}$ divided by $|\textbf{X}|$.
If the returned set $\mathcal{Q}$ is empty, we double $r$ and repeat \textsc{TreeSearch} until $\mathcal{Q}$ is non-empty.
Now, so long as $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| < k$, we repeat \textsc{TreeSearch}, but instead of doubling $r$, we multiply it by a factor determined by the LFD in the vicinity of the query ball.
In particular, we increase the radius by a factor of $\big( {\frac{k}{\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}|}} \big)^\frac{1}{\mu}$,
where $\mu$ is the harmonic mean of the LFD of the clusters in $\mathcal{Q}$, to ensure that this factor is not dominated by outlier clusters with very high LFD.
Intuitively, this factor will be large when we have found few points or the LFD is low, because in these cases, we can afford to increase the radius quickly. It will be small when we have found almost all $k$ points or LFD is large, and thus we want to increase the radius slowly.
Once $\sum_{\mathcal{C} \in \mathcal{Q}} |\mathcal{C}| \geq k$, we perform \textsc{LeafSearch} (Alg~\ref{alg:methods:rnn-search:leaf-search}) to find the $k$ nearest neighbors.
\vspace{-0.1in}
\begin{algorithm}[h]
    \caption{\textsc{TreeSearch}($\mathcal{C}$, $q$, $r$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{algorithmic}
        \Require $\mathcal{C}$, a cluster
        \Require $q$, a query
        \Require $r$, a search radius
        \If{$\delta^+_\mathcal{C} \leq r$}
            \State \textbf{return} $\{\mathcal{C}\}$
        \EndIf
        \State $[\mathcal{L}, \mathcal{R}]$ $\Leftarrow$ \textit{children} of $\mathcal{C}$
        \State \textbf{return} \Call{TreeSearch}{$\mathcal{L}, q, r$} $\cup$ \Call{TreeSearch}{$\mathcal{R}, q, r$}
    \end{algorithmic}
\end{algorithm}
\vspace{-0.2in}

@@

In CHESS~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(\textbf{X})}^{\textrm{metric entropy}}
        }_{\textrm{\textsc{TreeSearch}}}
        +
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ {\text{LFD}(q, \rho)}}^{\textrm{scaling factor}}
        }_{\textrm{\textsc{LeafSearch}}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\hat{r}$ is the mean cluster radius, $\mathcal{N}_{\hat{r}}(\textbf{X})$ is the minimum number of clusters of radius $\hat{r}$ needed to cover $\textbf{X}$.

### 2025-11-15 sections/3-datasets.tex
brought back some stuff

    \caption{Local fractal dimension vs. cluster depth across six datasets; note the different y-axes. The x-axis is tree depth; the y-axis is the LFD of clusters at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max. The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD, indicating a manifold structure. The Random dataset has high LFD, indicating lack of manifold structure. The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process. It exhibits behavior akin to the Random dataset across three separate radial scales, as indicated by the three peaks in LFD at different depths in the tree.}
    \label{fig:results:lfd-plots}
\end{figure}

>>>>

    \caption{Local fractal dimension vs. cluster depth across six datasets;
    note the different y-axes.
    The x-axis is tree depth;
    the y-axis is the LFD of clusters at that depth.
    We plot the 5th, 25th, 50th, 75th, and 95th percentiles, and min/max.
    The Fashion-MNIST, Glove-25, Sift and Silva datasets are real-world datasets and exhibit low LFD, indicating a manifold structure.
    The Random dataset has high LFD, indicating lack of manifold structure.
    The measured LFD of ``only'' 20 is a consequence of Equation~\ref{eq:methods:lfd-original} and only having sampled $10^6$ points in a 128-dimensional space $(\log_2 10^6 \approx 20)$;
    We would need to sample many more points ($>2^{128}$), for the LFD to approach the embedding dimension of 128.
    The RadioML dataset is also a randomly generated dataset, but with a much more elaborate generation process.
    It exhibits behavior akin to the Random dataset across three separate radial scales, as indicated by the distinct peaks in LFD at different depths.}
    \label{fig:results:lfd-plots}
\end{figure}
